---
title: 'Machine Learning Project: Potability'
author: "Quentin Le Boulch and Nico Palermo"
date: "2024-05-18"
output:
  html_document:
    theme: cerulean
    toc: true
    toc_float: true
  pdf_document:
    toc: true
---
## Introduction

This project aims to analyze water potability using various machine learning techniques applied to a specific dataset. We set up several models to evaluate the prediction capacity of each method and compare their performances.
We'll see several supervised classification techniques by training set model, tree, random forrest
logit model and tree nnet models




## Library

We installed several library to perform our different test and models
```{r include=FALSE}

library(readxl)
library(summarytools)
library(corrplot)
library(caret)
library(dplyr)
library(lattice)
library(ggplot2)
library(FactoMineR)
library(factoextra)
library(readxl)
library(MASS)
library(caTools)
library(ggplot2)
library(caret)
library(nnet)
library(rpart)
library(rpart.plot) 
library(tree) 
library(randomForest)
library(gbm)


```




## 1. Dataset

We download the dataset from Kaggle: https://www.kaggle.com/datasets/nayanack/water-probability/data

The data provides information about whether the water is potable or not, based on various water characteristics variables.

There is 9 explicative variables and the 10th variable is whether the is potable or not:

pH: pH of the water (measured in pH units).

Hardness: Hardness of the water (measured in mg/L).

Solids: Total dissolved solids in the water (measured in ppm).

Chloramines: Amount of chloramines in the water (measured in ppm).

Sulfate: Amount of sulfate in the water (measured in mg/L).

Conductivity: Conductivity of the water (measured in μS/cm).

Organic_carbon: Amount of organic carbon in the water (measured in ppm).

Trihalomethanes: Amount of trihalomethanes in the water (measured in μg/L).

Turbidity: Turbidity of the water (measured in NTU).

Potability: Potability of the water (1 indicates potable, 0 indicates non-potable).

Summary Statistics:

The dataset contains 3276 entries.
The columns "pH", "Sulfate", and "Trihalomethanes" have missing values.
The data types of the columns are float64 for features and int64 for the target variable (Potability).

```{r}
setwd("C:/Users/quent/OneDrive - Université Côte d'Azur/Documents/MasterEE/s2/Big Data")



ptbl<- read.csv("water_potability.csv")

str(ptbl)
head(ptbl)
summary(ptbl)
```

## 2 Training Model

We create the training model with 3/4 of our data
We create the test set to test our model. The test set is composed by the
individual who aren't in the training set so 25%

There are 2457 on the train and 819 in testing


```{r}

ptbl_train <- createDataPartition(y=ptbl$Potability, p=0.75, times=1, list=FALSE)
training <- ptbl[ptbl_train,]
dim(training)
testing <- ptbl[-ptbl_train,]
dim(testing)

```



We stacked the data and created the plots relating all the variables we have

```{r}
summary(training)


featurePlot(x=training, y = training$Potability,)

```



## 3 Tree 


The use of a decision tree in our project serves to better understand the data, identifying important variables

```{r}
tree <- rpart(Potability~.,data=training, method = "class")
prp(tree,extra=1,faclen=5, box.col=c("red","green")[tree$frame$yval])
```


This tree give us a lot of informations. In fact we can see our model follow 
specific conditions:
1st condition depending on sulfate: if sulfate <259 the water is prédicted 
not patable with 49 who are really not patable and 17 who are potable so with wrong prediction.
If sulfate >= 259 then we go to next condition who is Sulfate <409. And we follow
the tree to know if the water potability prédiction détermined by the tree structure
Based on the distribution of the tree we can think Sulfate is the variable who
is the most importante.

To verify this supposition we check forvariables importance on the tree


```{r}
tree$variable.importance
barplot(tree$variable.importance, horiz = TRUE, col = "purple")
```


This graph give the importance of each variable. We can see the most importante is Sulfate like our assumption. 
We also can see Sulfate importance =37 and if we add all the other variables importance we are near to 37 so we can said Sulfate determine for a very big part of our model like approximately 50% of the model. 
We can also see on the graph Turdibity for example as a very small impact on our model and we we could expect this because the variable is not even in the tree






This graph is the comparison between the variable ph and Sulfate because they are
the most impactful variable to determine if the water is potable or not

```{r}
qplot(ph, Sulfate, data=training)

```



Next, we test correlation between the 2 variables
```{r}
 
cor.test(ptbl$ph,ptbl$Sulfate)

```

The results show that the correlation is not significant (p-value is large 0.4026).




We compute the prediction with our data test and after that we're able to compute 
the confusion matrix. Confusion matrix allow to asses the precision and accuracy of 
our supervised classification.

```{r}

tree_prediction <- predict(tree, testing, type="class")
confusionMatrix(as.factor(testing$Potability),tree_prediction)
```


According to the confusion matrix, there were 502 correct predictions and 317
wrong prediction. The accuracy is about 61.29%

The true Negative Rate (TNR), specificity:
TN/(TN + FN) = 80/(80 +73) = 0.5229

The True Postive Rate (TPR), sensitivity
TP/(TP + FP) = 422/(422 + 244) = 0.6336

The accuracy:
TP + TN/ TP +TN +FP +FN = 0.6129 



## 4 Random Forest 


The random forest algorithm creates multiple decision trees using different subsets of the data and averages their predictions to improve accuracy and prevent overfitting. It highlights the most important variables contributing to the model's predictions by assessing their impact across all trees.


For this method, we train it on training, then we predict thanks to this training 
on testing and we compare.


```{r}

training_complete <- na.omit(training)
randomf1 <- randomForest(as.factor(training_complete$Potability) ~ ., data = training_complete)
randomf1_pred <- predict(randomf1, testing, type="response")

importance(randomf1)
```

A higher value means that the feature contributes more to reducing uncertainty
about class labels, and is therefore more important.
So this give usthe importance of each variable in the creation of the tree
Sulfate and PH have the mean decrease Gini higher so there are the most importante variables.




```{r}
confusionMatrix(as.factor(testing$Potability),randomf1_pred)

plot(c(1:819), testing$Potability, col ="green")

```
We estimated a model with a better accuracy than the previous one (69%).
The sensitivity and specificity is narrow the same.
 
The graph give us the distribution of the prediction of
the variable potability for our group test.



## 5. Logit Model

We did another model, the logit one. The logit model, is a supervised machine learning algorithm used for binary classification tasks. It estimates the probability that a given input belongs to a certain class by applying the logistic function to a linear combination of the input features.


```{r include=FALSE}
logit_model <- glm(Potability ~ ., data=training, family=binomial(link="logit"))
predictions <- predict(logit_model, testing, type = "response")
predictions
```

```{r}
datapr = as.numeric(predictions>0.75)
reference = testing$Potability
confusionMatrix(as.factor(datapr), as.factor(reference))

```

The model predict 0 for each result so all the water to be . This model is not 
good because we predict for all individuals 0. We can see the accuracy is very low 59%
Maybe something is wrong ? We dont take this model as a reference to explain our study.



## 6. Tree Nnet

The TreeNet model combines decision trees and neural networks to leverage the strengths of both techniques. It uses decision trees to handle non-linear relationships and interactions, while neural networks refine these predictions for improved accuracy in classification tasks.

```{r include=FALSE}
treeNnet <- nnet(Potability ~ ., data=training, size=2)

```

```{r include=FALSE}
nn_prob <- predict(treeNnet, testing, type="raw")
nn_prob
```

```{r}
data = as.numeric(nn_prob>0.5)
reference = testing$Potability
confusionMatrix(as.factor(data), as.factor(reference))
```

The model predict 0 for each result so all the water to be . This model is not 
good because we predict for all individuals 0. We can see the accuracy is very low 58.73%



##7 A new to asses mising values

For now we're testing a model with droping all observations where there is missing values.
Indeed, our approach involve dropping observations with missing values, have reduced the dataset's overall size and affected model performance. More sophisticated imputation techniques could yield better results

```{r}
training2 <- na.omit(training)
testing2 <- na.omit(testing)
```
 
We did it again a tree and a confusion matrix...
 
```{r}
tree <- rpart(Potability~.,data=training2, method = "class")
prp(tree,extra=1,faclen=5, box.col=c("red","green")[tree$frame$yval])
```


```{r}
tree_prediction2 <- predict(tree, testing, type="class")
confusionMatrix(as.factor(testing$Potability),tree_prediction2)

```

This model gave us  a bit lower accuracy than first one. 





## 8. Conclusion and limits


In summary, our study aimed to predict water potability using various machine learning models. We explored decision trees, random forests, logistic regression, and TreeNet models, identifying sulfate and pH as significant predictors.

Some models, like logistic regression and TreeNet, showed poor performance, indicating potential issues with model choice or parameter settings. Further tuning and optimization are required.
The study did not explore advanced feature machine learning techniques that could potentially improve the models predictive power.
The dataset's size and the presence of missing or noisy data may have impacted the accuracy and generalizability of our models. Also, our study did not explicitly address the presence of outliers in the dataset. Outliers can skew the results and affect the accuracy of the models, potentially leading to biased predictions. Detecting and appropriately handling outliers could improve the robustness and reliability of our models.














